---
title: "flexdashboard — Part 2: Adding Shiny"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: flatly
    navbar:
      - { title: "Run Document in RStudio — do not Knit", align: right }
runtime: shiny
---

```{r global, include=FALSE}
# Label this chunk "global" — it runs once in the global environment and its
# objects are shared across all user sessions. This is the official recommendation
# for loading data in Shiny flexdashboards (see the flexdashboard docs).

library(flexdashboard)
library(shiny)
library(ggplot2)
library(plotly)
library(DT)
library(dplyr)
library(scales)

# ── Load simulated IPD co-etiology data ────────────────────────────────────
# Run simulate_data.R once before opening this dashboard to create the file.
#   source("simulate_data.R")
ipd_data <- readRDS("data/ipd_sim.Rds")

# ── Constants ───────────────────────────────────────────────────────────────
PATHOGENS <- c(
  "rotavirus", "norovirus", "adenovirus", "astrovirus", "sapovirus",
  "st_etec", "shigella", "campylobacter", "tepec",
  "v_cholerae", "salmonella", "cryptosporidium"
)

pathogen_labels <- c(
  rotavirus       = "Rotavirus",
  norovirus       = "Norovirus GII",
  adenovirus      = "Adenovirus",
  astrovirus      = "Astrovirus",
  sapovirus       = "Sapovirus",
  st_etec         = "ST-ETEC",
  shigella        = "Shigella",
  campylobacter   = "C. jejuni/coli",
  tepec           = "tEPEC",
  v_cholerae      = "V. cholerae",
  salmonella      = "Salmonella",
  cryptosporidium = "Cryptosporidium"
)

STUDIES <- c("ABCD", "EFGH", "GEMS", "VIDA", "MALED")

study_colors <- c(
  ABCD  = "#00468B",
  EFGH  = "#925E9F",
  GEMS  = "#42B540",
  VIDA  = "#0099B4",
  MALED = "#ED0000"
)

# ── Shared helper — the filter_base() pattern from the real dashboard ───────
# A plain R function: easy to read, test, and reuse across tabs.
# Accepts plain arguments (not reactive values) — called inside reactive().
filter_base <- function(data, msd_only, study_sel) {
  d <- data
  if (msd_only)           d <- filter(d, MSD == 1)
  if (length(study_sel))  d <- filter(d, study %in% study_sel)
  d
}

# Shared ggplot theme — mirrors base_theme() in the real dashboard
base_theme <- function() {
  theme_minimal(base_size = 13) +
    theme(
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
}
```


Why Shiny? {data-icon="fa-bolt"}
=====================================

Row {data-height=100}
-------------------------------------

### Moving from static to reactive

Part 1 showed what you can do without Shiny: ggplot figures, plotly hover, and DT filters
all embedded in a static HTML file. This tab explains what Shiny adds, when you would need it, and what the trade-offs are.

Row {data-height=350}
-------------------------------------

### What the static dashboard cannot do

The plotly and DT interactivity in Part 1 let users interact with results that were already computed when you knit the dashboard. **Shiny keeps R running**, so an input can trigger new calculations and update multiple figures or tables at once. For example, users could:

- Filter the data by site, population, or date range
- Choose an outcome and recalculate a summary
- Change an analysis setting and redraw several figures
- Make one selection that updates plots and tables across multiple tabs

Adding `runtime: shiny` lets the dashboard respond to those choices while it runs.

### {data-width=180}

```{r}
gif_path <- "figures/cluster_fire.gif"
if (file.exists(gif_path)) {
  knitr::asis_output(
    paste0('<img src="', knitr::image_uri(gif_path),
           '" style="max-width:100%;max-height:180px;display:block;margin:auto;">')
  )
} else {
  cat("*Place cluster_fire.gif in the figures/ folder*")
}
```

### What Shiny adds: one line in the YAML

```yaml
title: "flexdashboard — Part 2: Adding Shiny"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: flatly
    navbar:
      - { title: "Run Document in RStudio — do not Knit", align: right }
runtime: shiny # <----- add this line
```

By specifying **runtime: shiny**, the dashboard is no longer a document that knits to a file.
It is an **interactive document** that needs a live R session to run. R stays open,
listens for user input, and re-executes code when inputs change.

Row {data-height=400}
-------------------------------------

### The trade-off: file vs. application

| Static flexdashboard (Part 1) | Shiny flexdashboard (Part 2) |
|---|---|
| Knit → single `.html` file | Run Document → live application |
| No server needed to open | Needs R running: locally or deployed |
| Email as an attachment | Share a URL (Posit Connect) |
| All data baked into the file | Data stays on the server |
| Fixed at knit time | Re-runs R code when inputs change |
| Fine for small/medium data | Better for large or sensitive data |
| Data visible in page source | Data never leaves the server |

### How to run and deploy

**Locally** — click **Run Document** in RStudio (not Knit). The dashboard opens in a
browser window and stays live while your RStudio session is open.

**Posit Connect Cloud** — is the recommended platform for new deployments. To publish from the R console:

```r
install.packages("rsconnect")       # once
rsconnect::connectCloudUser()        # once; opens a browser to authorize access
rsconnect::deployApp()               # run from the dashboard's folder
```

You can also publish using the Publish button in RStudio. Connect Cloud has a free plan, but content on that plan is public; use an approved private hosting option for sensitive data. 

Shiny Concepts {data-icon="fa-cogs"}
=====================================

Column {data-width=350}
-------------------------------------

### 1. The user makes a selection

Shiny inputs are controls such as dropdowns, checkboxes, and sliders. Each has
an **ID** that you use to read its current value in R.

```r
selectInput("study",    # input id
            "Study:",   # label
            choices = c("Study A", "Study B", "Study C"))
```

If the user selects Study B, `input$study` has the value `"Study B"`.

Column {data-width=350}
-------------------------------------

### 2. R prepares the selected data

`reactive()` defines a calculation that can change when its inputs change.
Here, `selected_data()` means “filter the data to the currently selected study.”

```r
selected_data <- reactive({
  data |>
    dplyr::filter(study == input$study)
})
```

Shiny sees that this calculation depends on `input$study`. If the user selects
another study, Shiny marks the old result as out of date and calculates the new
one when it is needed.

The `()` matter: **`selected_data()` gets the current result** of the reactive
calculation.

Column {data-width=400}
-------------------------------------

### 3. The dashboard displays a result

A `render*()` function creates an output. This plot uses the currently
selected data:

```r
renderPlot({
  ggplot(selected_data(), aes(x = age)) +
    geom_histogram()
})
```

When `input$study` changes, `selected_data()` changes, so Shiny redraws this
plot. A table or count that also uses `selected_data()` would update too.

| Function | Creates |
|---|---|
| `renderPlot()` | R plot, including ggplot |
| `renderPlotly()` | Interactive Plotly plot |
| `renderDT()` | DT table |
| `renderText()` | Updating text |

Shiny follows the dependencies in the code: changing an unrelated input
does not recalculate `selected_data()`.

Example 1: Pathogen Prevalence {data-icon="fa-chart-bar"}
=====================================

Column {.sidebar}
-------------------------------------

### Filters

```{r}
checkboxInput("pprev_msd_only",
              "MSD cases only",
              value = TRUE)

checkboxGroupInput("pprev_studies",
                   "Studies:",
                   choices  = STUDIES,
                   selected = STUDIES,
                   inline   = TRUE)
```

This tab displays **Pathogen Prevalence**. The sidebar
(`{.sidebar}` column) contains inputs on the left, with output
filling the remaining width of the page. Change either filter and watch the chart update.

Row {data-height=50}
-------------------------------------

### Attributable prevalence — change the filters on the left

Row {data-height=600}
-------------------------------------

### Prevalence chart

```{r}
pprev_data <- reactive({
  filter_base(ipd_data, input$pprev_msd_only, input$pprev_studies)
})

renderPlotly({
  d <- pprev_data()
  req(nrow(d) > 0)

  prev_df <- data.frame(
    pathogen = PATHOGENS,
    label    = pathogen_labels[PATHOGENS],
    n_total  = nrow(d),
    stringsAsFactors = FALSE
  ) %>%
    mutate(
      n_attr = sapply(PATHOGENS, function(p)
                      sum(d[[paste0(p, "_tac_attr")]] == 1, na.rm = TRUE)),
      rate   = n_attr / n_total
    ) %>%
    arrange(desc(rate)) %>%
    mutate(label = factor(label, levels = rev(label)))

  p <- ggplot(prev_df,
              aes(x    = label,
                  y    = rate,
                  text = paste0(label,
                                "\n", percent(rate, accuracy = 0.1),
                                "  (n = ", n_attr, " / ", n_total, ")"))) +
    geom_col(fill = "#08306B", alpha = 0.80, width = 0.65) +
    geom_text(aes(label = percent(rate, accuracy = 0.1)),
              hjust = -0.1, size = 3.5, color = "gray20") +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.15))) +
    coord_flip() +
    labs(x        = NULL,
         y        = "Attributable prevalence",
         subtitle = paste0("n = ", nrow(d))) +
    base_theme() +
    theme(panel.grid.major.y = element_blank())

  ggplotly(p, tooltip = "text") %>%
    layout(margin = list(l = 150))
})
```

### Shiny controls in this example

**Sidebar layout:** `{.sidebar}` puts the controls beside the chart.

```markdown
Column {.sidebar}
-------------------------------------
```

**One on/off choice:** `checkboxInput()` returns `TRUE` or `FALSE`.

```r
checkboxInput("pprev_msd_only", "MSD cases only", value = TRUE)
# Read its value with input$pprev_msd_only
```

**Several choices:** `checkboxGroupInput()` returns a vector of the selected
studies.

```r
checkboxGroupInput("pprev_studies", "Studies:",
                   choices = STUDIES, selected = STUDIES)
# Read the selections with input$pprev_studies
```

The selected values feed into `pprev_data()`, which the chart uses:

```r
pprev_data <- reactive({
  filter_base(ipd_data, input$pprev_msd_only, input$pprev_studies)
})

renderPlotly({
  d <- pprev_data()
  # Build the chart using d
})
```


Example 2: Co-infection Prevalence {data-icon="fa-chart-bar"}
=====================================

Row {data-height=80}
-------------------------------------

### Among cases attributable to one pathogen, how often are others present?

Row {data-height=200}
-------------------------------------

### Controls {data-width=320}

```{r}
inputPanel(
  selectInput("coinf_index", "Primary pathogen:",
              choices  = setNames(PATHOGENS, pathogen_labels),
              selected = "shigella"),

  radioButtons("coinf_def", "Co-infection definition:",
               choices  = c("Attributable" = "tac_attr",
                            "Any detection" = "bin"),
               selected = "tac_attr",
               inline   = TRUE),

  checkboxInput("coinf_msd_only", "MSD cases only", value = TRUE),

  checkboxGroupInput("coinf_studies", "Studies:",
                     choices  = STUDIES,
                     selected = STUDIES,
                     inline   = TRUE)
)
```

Row {data-height=800}
-------------------------------------

### Co-infection rates among index-pathogen cases {data-width=600}

```{r}
index_cases <- reactive({
  d   <- filter_base(ipd_data, input$coinf_msd_only, input$coinf_studies)
  col <- paste0(input$coinf_index, "_tac_attr")
  req(col %in% names(d))
  filter(d, .data[[col]] == 1)
})

renderPlotly({
  d       <- index_cases()
  req(nrow(d) > 0)

  ip      <- input$coinf_index
  copaths <- setdiff(PATHOGENS, ip)
  suffix  <- input$coinf_def        # "tac_attr" or "bin"

  coinf_df <- data.frame(
    pathogen = copaths,
    label    = pathogen_labels[copaths],
    n_index  = nrow(d),
    stringsAsFactors = FALSE
  ) %>%
    mutate(
      n    = sapply(copaths, function(cp)
                   sum(d[[paste0(cp, "_", suffix)]] == 1, na.rm = TRUE)),
      rate = n / n_index
    ) %>%
    arrange(rate) %>%
    mutate(label = factor(label, levels = label))

  p <- ggplot(coinf_df,
              aes(x    = label,
                  y    = rate,
                  text = paste0(label,
                                "\n", percent(rate, accuracy = 0.1),
                                "  (n = ", n, " / ", n_index, ")"))) +
    geom_col(fill = "#6BAED6", alpha = 0.85, width = 0.65) +
    geom_text(aes(label = percent(rate, accuracy = 0.1)),
              hjust = -0.1, size = 3.5) +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.15))) +
    coord_flip() +
    labs(x        = NULL,
         y        = "Co-infection rate",
         subtitle = paste0("Among attributable ", pathogen_labels[ip],
                           " cases  |  n = ", nrow(d))) +
    base_theme() +
    theme(panel.grid.major.y = element_blank())

  ggplotly(p, tooltip = "text") %>%
    layout(margin = list(l = 150))
})
```

### Shiny controls in this example

**Dropdown:** `selectInput()` lets the user choose one primary pathogen.

```r
selectInput("coinf_index", "Primary pathogen:",
            choices = setNames(PATHOGENS, pathogen_labels),
            selected = "shigella")
# Selected value: input$coinf_index
```

**Radio buttons:** `radioButtons()` lets the user choose one definition. The
labels appear on the page; `"tac_attr"` or `"bin"` is passed to R.

```r
radioButtons("coinf_def", "Co-infection definition:",
             choices = c("Attributable" = "tac_attr",
                         "Any detection" = "bin"))
# Selected value: input$coinf_def
```

These selections affect different parts of the chart code:

```r
index_cases <- reactive({
  # Uses input$coinf_index to choose the primary pathogen's cases
})

renderPlotly({
  d <- index_cases()
  suffix <- input$coinf_def
  # Uses suffix to choose which co-infection rates to plot
})
```

Changing the primary pathogen updates `index_cases()` and the chart. Changing
the definition updates the chart without needing to recalculate `index_cases()`.


Example 3: Age Distribution {data-icon="fa-chart-area"}
=====================================

Row {data-height=80}
-------------------------------------

### How does age at presentation vary by attribution status? — mirrors the real Age Distributions tab

Row {data-height=200}
-------------------------------------

### Controls {data-width=320}

```{r}
inputPanel(
  selectInput("age_pathogen", "Pathogen:",
              choices  = setNames(PATHOGENS, pathogen_labels),
              selected = "shigella"),

  checkboxInput("age_msd_only", "MSD cases only", value = TRUE),

  checkboxGroupInput("age_studies", "Studies:",
                     choices  = STUDIES,
                     selected = STUDIES,
                     inline   = TRUE),

  checkboxInput("age_show_attr",    "Attributable",               value = TRUE),
  checkboxInput("age_show_nonattr", "Detected (not attributable)", value = TRUE),
  checkboxInput("age_show_nodet",   "Not detected",                value = FALSE),

  sliderInput("age_bw", "Density bandwidth:",
              min = 0.3, max = 3, value = 1, step = 0.1)
)
```

Row {data-height=525}
-------------------------------------

### Age density by attribution status

```{r}
# age_subset() classifies cases and is shared by BOTH plots on this tab.
# The pathogen dropdown triggers both plots to update together.
age_subset <- reactive({
  d  <- filter_base(ipd_data, input$age_msd_only, input$age_studies)
  ip <- input$age_pathogen
  ac <- paste0(ip, "_tac_attr")
  dc <- paste0(ip, "_bin")
  req(all(c(ac, dc) %in% names(d)))

  d %>%
    mutate(age_group = case_when(
      .data[[ac]] == 1                    ~ "Attributable",
      .data[[dc]] == 1 & .data[[ac]] == 0 ~ "Detected (not attributable)",
      TRUE                                ~ "Not detected"
    ))
})

renderPlotly({
  d <- age_subset()

  groups_show <- c(
    if (input$age_show_attr)    "Attributable",
    if (input$age_show_nonattr) "Detected (not attributable)",
    if (input$age_show_nodet)   "Not detected"
  )
  d <- filter(d, age_group %in% groups_show)
  req(nrow(d) > 0)

  attr_colors <- c(
    "Attributable"               = "#00468B",
    "Detected (not attributable)"= "#ED0000",
    "Not detected"               = "#ADB6B6"
  )

  p <- ggplot(d, aes(x = age, fill = age_group, color = age_group)) +
    geom_density(alpha = 0.45, adjust = input$age_bw, linewidth = 0.8) +
    scale_fill_manual(values  = attr_colors) +
    scale_color_manual(values = attr_colors) +
    scale_x_continuous(breaks = seq(0, 60, 6), limits = c(0, 60)) +
    labs(x        = "Age (months)",
         y        = "Density",
         fill     = NULL, color = NULL,
         subtitle = pathogen_labels[input$age_pathogen]) +
    base_theme()

  ggplotly(p) %>%
    layout(legend = list(orientation = "h", y = -0.2))
})
```

### Shiny controls in this example

**Checkboxes can control what appears in a plot.** Each checkbox has its own
`TRUE` or `FALSE` value:

```r
checkboxInput("age_show_attr", "Attributable", value = TRUE)
checkboxInput("age_show_nonattr", "Detected (not attributable)", value = TRUE)
checkboxInput("age_show_nodet", "Not detected", value = FALSE)
```

The density plot reads those values to decide which groups to display:

```r
groups_show <- c(
  if (input$age_show_attr)    "Attributable",
  if (input$age_show_nonattr) "Detected (not attributable)",
  if (input$age_show_nodet)   "Not detected"
)
```

**Slider:** `sliderInput()` supplies a number that the plot can use directly.

```r
sliderInput("age_bw", "Density bandwidth:",
            min = 0.3, max = 3, value = 1, step = 0.1)

geom_density(adjust = input$age_bw)
```

The pathogen and study selections are used by `age_subset()`, which supplies
data to both figures. The group checkboxes and bandwidth slider are used only
by the density plot.

Row {data-height=480}
-------------------------------------

### Proportion attributable by 6-month age bin

```{r}
renderPlotly({
  d <- age_subset()

  attr_colors <- c(
    "Attributable"               = "#00468B",
    "Detected (not attributable)"= "#ED0000",
    "Not detected"               = "#ADB6B6"
  )

  bin_df <- d %>%
    mutate(age_bin = cut(age, breaks = seq(0, 60, 6),
                         include.lowest = TRUE, right = FALSE)) %>%
    filter(!is.na(age_bin)) %>%
    group_by(age_bin, age_group) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(age_bin) %>%
    mutate(pct = n / sum(n))

  p <- ggplot(bin_df,
              aes(x    = age_bin,
                  y    = pct,
                  fill = age_group,
                  text = paste0(age_group, ": ",
                                percent(pct, accuracy = 0.1),
                                "  (n = ", n, ")"))) +
    geom_col(position = "fill", width = 0.85) +
    scale_y_continuous(labels = percent_format()) +
    scale_fill_manual(values = attr_colors) +
    labs(x    = "Age bin (months)",
         y    = "Proportion",
         fill = NULL) +
    base_theme() +
    theme(axis.text.x = element_text(angle = 35, hjust = 1))

  ggplotly(p, tooltip = "text") %>%
    layout(legend = list(orientation = "h", y = -0.3))
})
```

