---
title: "Build Your Own Dashboard — Practice"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
    theme: flatly
runtime: shiny
---

```{r global, include=FALSE}
library(flexdashboard)
library(shiny)
library(ggplot2)
library(plotly)
library(DT)
library(dplyr)

# Run source("simulate_data.R") once before opening this dashboard.
# It creates the data file used in both lessons and this exercise.
my_data <- readRDS("data/ipd_sim.Rds")
my_data <- my_data |> mutate(study = as.character(study))
```

Start Here {data-icon="fa-play"}
=====================================

Column
-------------------------------------

### Your task

Build a small dashboard using the tools from Parts 1 and 2. The file supplies
**data and a page layout**; you write the plots, table, controls, and reactive
code. The practice dataset is simulated and has `study`, `age`, `enr_haz`,
and `MSD` (0/1) columns.

1. In the R console, run `source("simulate_data.R")` to create `data/ipd_sim.Rds`.
2. Open this `.Rmd` in the same project folder as the lessons.
3. Complete the `TODO` chunks in order, clicking **Run Document** as you go.
   This is a Shiny document, so use **Run Document** throughout.

The first two pages cover flexdashboard, ggplot, Plotly, and DT. The last page
adds Shiny to update results when someone uses a control. You can replace the
simulated data with your own after the practice dashboard works.

### What you will build

- A figure of age versus baseline HAZ, arranged on a dashboard page
- An interactive version with hover labels and a searchable summary table
- A plot and record count that respond to the same Shiny controls


1. Make a dashboard page {data-icon="fa-chart-area"}
=====================================

Column {data-width=600}
-------------------------------------

### Age and baseline HAZ

**TODO 1:** Make a `ggplot` of `enr_haz` versus `age`. Color the points by
`study`, add axis labels, and print the figure. This is ordinary R Markdown
code inside a flexdashboard box.

```{r}
# Start here:
# p <- ggplot(my_data, aes(x = age, y = enr_haz, color = study)) +
#   geom_point(...)
# p
```

Column {data-width=400}
-------------------------------------

### Tell the reader what the figure shows

Write one or two sentences here about the figure you made. In your own
research dashboard, you could put an interpretation or a data caveat beside
the chart.

**Layout challenge:** Adjust the `data-width` values of the two columns above.
What changes when you run the document again? Try adding a second `###` box
below this text.


2. Add browser interactivity {data-icon="fa-mouse-pointer"}
=====================================

Column {data-width=600}
-------------------------------------

### Plotly figure

**TODO 2:** Make an interactive version of your figure. Add a `text` aesthetic
that shows the study, age, and HAZ on hover. Then call
`ggplotly(p, tooltip = "text")`. You can copy and adapt the ggplot from TODO 1.

```{r}
# p <- ggplot(my_data, aes(x = age, y = enr_haz, color = study,
#                          text = paste0(...))) +
#   geom_point(...)
# ggplotly(p, tooltip = "text")
```

Column {data-width=400}
-------------------------------------

### Searchable study table

**TODO 3:** Calculate the number of records in each study, then display your
summary with `DT::datatable()`. Try searching for a study and sorting by count.

```{r}
# study_summary <- my_data |>
#   group_by(study) |>
#   summarise(n = n(), .groups = "drop")
# datatable(study_summary, rownames = FALSE)
```

These Plotly and DT interactions work in the browser without a Shiny input.


3. Add Shiny controls {data-icon="fa-bolt"}
=====================================

Column {.sidebar}
-------------------------------------

### Select the cases

**TODO 4:** Add two controls here:

- `selectInput("practice_study", ...)` to choose one study or "All studies"
- `checkboxInput("practice_msd", ...)` to turn "MSD cases only" on or off

Use `sort(unique(my_data$study))` for the study choices. Put the controls in
this sidebar so they appear beside the results.

```{r}
# Write your selectInput() and checkboxInput() here.
```

Column
-------------------------------------

### Results for the current selection

**TODO 5:** Complete `selected_data` so it uses `input$practice_study` and
`input$practice_msd` to select rows. When "All studies" is selected, keep all
studies. When the checkbox is checked, keep rows with `MSD == 1`.

```{r}
selected_data <- reactive({
  d <- my_data
  # Add the study and MSD selection steps here.
  d
})
```

**TODO 6:** Create two outputs that call `selected_data()`:

- A `renderPlotly({ ... })` plot of age versus HAZ for the selected cases
- A `renderText({ ... })` count using `nrow(selected_data())`

You can adapt your Plotly code from TODO 2. Use `req(nrow(d) > 0)` before
plotting so an empty selection does not cause an error.

```{r}
# renderPlotly({
#   d <- selected_data()
#   req(nrow(d) > 0)
#   p <- ggplot(d, aes(...)) + ...
#   ggplotly(p)
# })

# renderText({
#   paste("Records in current selection:", nrow(selected_data()))
# })
```

### Check your dashboard

Change the study, then toggle "MSD cases only." Do **both** the plot and the
record count update? If so, both outputs are reading the same reactive result.

**If you finish early:** Add a `sliderInput()` for an age range and use its
value inside `selected_data()`. The two outputs should update without changing
their code.

