---
title: "BYOD: Build Your Own Dashboard"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: flatly
    navbar:
      - { title: "Replace the TODOs — then make it yours", align: right }
runtime: shiny
---

```{r global, include=FALSE}
# ── Packages ───────────────────────────────────────────────────────────────
library(flexdashboard)
library(shiny)
library(ggplot2)
library(plotly)
library(DT)
library(dplyr)
library(scales)

# ══════════════════════════════════════════════════════════════════════════
# TODO 0 — LOAD YOUR DATA
#
# The placeholder loads the simulated IPD co-etiology dataset created by
# simulate_data.R. Run that script once before opening this dashboard:
#   source("simulate_data.R")
#
# To swap in your own data:
#   1. Replace the readRDS() line with your own file path
#   2. Keep the name my_data
#   3. Update PATHOGENS, pathogen_labels, and STUDIES below to match
# ══════════════════════════════════════════════════════════════════════════
my_data <- readRDS("data/ipd_sim.Rds")

# ── Constants — update these if using your own data ─────────────────────────
PATHOGENS <- c(
  "rotavirus", "norovirus", "adenovirus", "astrovirus", "sapovirus",
  "st_etec", "shigella", "campylobacter", "tepec",
  "v_cholerae", "salmonella", "cryptosporidium"
)

pathogen_labels <- c(
  rotavirus       = "Rotavirus",
  norovirus       = "Norovirus GII",
  adenovirus      = "Adenovirus",
  astrovirus      = "Astrovirus",
  sapovirus       = "Sapovirus",
  st_etec         = "ST-ETEC",
  shigella        = "Shigella",
  campylobacter   = "C. jejuni/coli",
  tepec           = "tEPEC",
  v_cholerae      = "V. cholerae",
  salmonella      = "Salmonella",
  cryptosporidium = "Cryptosporidium"
)

STUDIES <- c("ABCD", "EFGH", "GEMS", "VIDA", "MALED")

# ── Shared filter helper ────────────────────────────────────────────────────
filter_base <- function(data, msd_only, study_sel) {
  d <- data
  if (msd_only)           d <- filter(d, MSD == 1)
  if (length(study_sel))  d <- filter(d, study %in% study_sel)
  d
}

# ══════════════════════════════════════════════════════════════════════════
# TODO 0b — UPDATE IF USING YOUR OWN DATA
# ══════════════════════════════════════════════════════════════════════════
numeric_cols     <- c("age", "enr_haz")
categorical_cols <- c("study", "MSD")
```


Instructions {data-icon="fa-info-circle"}
=====================================

Row {data-height=100}
-------------------------------------

### BYOD — Build Your Own Dashboard

Work through the tabs in order. Each tab has one exercise. Look for `# TODO` blocks
and `"TODO: ..."` placeholder text — replace them with real code and labels for
your data. Click **Run Document** after each change to see the result.

### {data-width=180}

```{r}
gif_path <- "figures/duck_dance.gif"
if (file.exists(gif_path)) {
  knitr::asis_output(
    paste0('<img src="', knitr::image_uri(gif_path),
           '" style="max-width:100%;max-height:160px;display:block;margin:auto;">')
  )
} else {
  cat("*Place duck_dance.gif in the figures/ folder*")
}
```

Row {data-height=280}
-------------------------------------

### How to work through this file

**Step 1 — Load your data (do this first)**

In the `global` chunk at the very top of the `.Rmd`, replace the `mtcars` block with your
own dataset. Name it `my_data`. Update `numeric_cols` and `categorical_cols` to match your
column names. Run Document to confirm it loads without errors.

**Step 2 — Work through the tabs**

Each tab is one exercise. The instructions are in the box to the right of the output.
Replace `# TODO` comments and `"TODO: ..."` strings with code that works for your data.

**Step 3 — Check your work**

Use the checklist on the last tab before considering the dashboard finished.

**Step 4 — Make it yours**

Once every exercise works, go back and customise. Change themes, labels, colors, and layout.
Add a new page tab. Adjust `data-height` values. This file is your starting point, not your
final product.

### Quick reference

```markdown
# Three layout levels:
Page Title {data-icon="fa-home"}   # ==== → navbar tab
=====================================

Row {data-height=300}              # ---- → row
-------------------------------------

### Box title                      # ### → panel
```

```r
# Shiny building blocks:
inputPanel(...)                    # group input controls
selectInput("id", "Label:", choices)
checkboxGroupInput("id", "Label:", choices)
sliderInput("id", "Label:", min, max, value)
radioButtons("id", "Label:", choices)

reactive({ ... })                  # cache shared filtered data
renderPlot({ ... })                # ggplot output
renderPlotly({ ... })              # plotly output
renderDT({ ... })                  # DT table output
req(condition)                     # stop silently if FALSE
```

Row {data-height=160}
-------------------------------------

### Your dataset

```{r}
renderPrint({
  cat("Rows:", nrow(my_data), "  Columns:", ncol(my_data), "\n\n")
  cat("Column names:\n")
  cat(paste(" ", names(my_data), collapse = "\n"), "\n\n")
  cat("First 3 rows:\n")
  print(head(my_data, 3))
})
```

### Notes about your data

Replace this text with notes about your dataset. You will thank yourself later.

- **Source:** *(where does this data come from?)*
- **One row = ** *(what does one observation represent?)*
- **Key outcome:** *(what are you trying to show?)*
- **Key grouping variable:** *(how do you want to split or filter?)*
- **Any cleaning done in the global chunk:** *(note it here)*


Exercise 1: valueBox {data-icon="fa-hashtag"}
=====================================

Row {data-height=80}
-------------------------------------

### Exercise 1 — Summarise your data with three valueBox outputs

A `valueBox` highlights a single number with a label and an optional icon. Put one per
`###` box inside a short row. The color signals meaning at a glance.

Row {data-height=130}
-------------------------------------

### KPI 1

```{r}
# ══════════════════════════════════════════════════════════════════════════
# TODO 1a — Replace value and caption with something real from my_data.
# Ideas: nrow(my_data), mean(my_data$column), sum(my_data$column == "X")
# ══════════════════════════════════════════════════════════════════════════
valueBox(
  value   = "???",
  caption = "TODO: short label",
  icon    = "fa-database",
  color   = "primary"
)
```

### KPI 2

```{r}
# TODO 1b
valueBox(
  value   = "???",
  caption = "TODO: short label",
  icon    = "fa-bar-chart",
  color   = "info"
)
```

### KPI 3

```{r}
# TODO 1c
valueBox(
  value   = "???",
  caption = "TODO: short label",
  icon    = "fa-check",
  color   = "success"
)
```

Row {data-height=280}
-------------------------------------

### Your notes for this exercise

- KPI 1 shows: *(replace with what you chose and why)*
- KPI 2 shows: *(replace)*
- KPI 3 shows: *(replace)*

### Reference: valueBox arguments

```r
valueBox(
  value   = nrow(my_data),       # any R expression: a number, string, or %
  caption = "Total records",     # short descriptive label below the value
  icon    = "fa-users",          # Font Awesome: fa-  | Ionicons: ion-
  color   = "primary"            # primary / success / info / warning / danger
                                 # or any hex color: "#2C7BB6"
)
```

**Worked example for `mtcars`:**

```r
# KPI 1 — number of cars in the dataset
valueBox(nrow(my_data), caption = "Cars", icon = "fa-car", color = "primary")

# KPI 2 — average miles per gallon, rounded to 1 decimal
valueBox(round(mean(my_data$mpg), 1), caption = "Mean MPG",
         icon = "fa-tachometer-alt", color = "info")

# KPI 3 — count of manual transmission cars
valueBox(sum(my_data$am == "Manual"), caption = "Manual transmission",
         icon = "fa-cog", color = "success")
```


Exercise 2: Static ggplot {data-icon="fa-chart-bar"}
=====================================

Row {data-height=80}
-------------------------------------

### Exercise 2 — Build a static ggplot that runs at render time (no Shiny inputs)

A plain ggplot in a code chunk works exactly as it does in any R Markdown document.
Write it, and it renders into the box when you click Run Document.

Row {data-height=440}
-------------------------------------

### Your static plot

```{r fig.height=4.5}
# ══════════════════════════════════════════════════════════════════════════
# TODO 2 — Replace x, y, color, and labels with your own variables.
# Add or change the geometry to suit your data.
# ══════════════════════════════════════════════════════════════════════════
ggplot(my_data, aes(x = wt, y = mpg, color = cyl)) +  # TODO: your variables
  geom_point(alpha = 0.7, size = 2.5) +               # TODO: try geom_boxplot, geom_histogram, geom_bar
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "TODO: write a descriptive title",
    subtitle = "TODO: one-line description of what this shows",
    x        = "TODO: x axis label with units",
    y        = "TODO: y axis label with units",
    color    = "TODO: legend title"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

### Geometry options

Choose the geometry that fits your data:

```r
# Two numeric variables — scatter plot
geom_point(alpha = 0.7, size = 2.5)

# One numeric, one categorical — box plot
ggplot(my_data, aes(x = cyl, y = mpg, fill = cyl)) +
  geom_boxplot(alpha = 0.8)

# One numeric variable — histogram
ggplot(my_data, aes(x = mpg, fill = am)) +
  geom_histogram(bins = 15, alpha = 0.7, position = "identity")

# One categorical variable — bar count
ggplot(my_data, aes(x = cyl, fill = am)) +
  geom_bar(position = "dodge")

# Split into panels by a grouping variable
+ facet_wrap(~ cyl, ncol = 3)
```

**Color scales:**

```r
scale_color_brewer(palette = "Set2")    # categorical, color aesthetic
scale_fill_brewer(palette  = "Blues")   # categorical, fill aesthetic
scale_color_viridis_c()                 # continuous color
```


Exercise 3: Plotly {data-icon="fa-mouse-pointer"}
=====================================

Row {data-height=80}
-------------------------------------

### Exercise 3 — Add hover tooltips by converting your ggplot to plotly

Take the ggplot you built in Exercise 2 (or a new one) and wrap it in `ggplotly()`.
Add a `text` aesthetic to control exactly what appears when the user hovers over a point.

Row {data-height=460}
-------------------------------------

### Your plotly chart

```{r}
# ══════════════════════════════════════════════════════════════════════════
# TODO 3 — Add a text aesthetic and wrap in ggplotly().
# Step 1: add text = paste0(...) inside aes()
# Step 2: wrap the ggplot object in ggplotly(p, tooltip = "text")
# ══════════════════════════════════════════════════════════════════════════
p <- ggplot(my_data,
            aes(x     = wt,           # TODO: your x variable
                y     = mpg,          # TODO: your y variable
                color = cyl,          # TODO: your color variable
                text  = paste0(       # TODO: build your tooltip
                  "Car: ",    car,    "\n",
                  "MPG: ",    mpg,    "\n",
                  "Weight: ", wt
                ))) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "TODO: title",
    x     = "TODO: x label",
    y     = "TODO: y label"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank())

ggplotly(p, tooltip = "text")
```

### How the tooltip works

The `text` aesthetic is what appears when the user hovers. Build it with `paste0()`:

```r
aes(text = paste0(
  "Record: ",  my_id,         "\n",
  "Value: ",   round(x, 1),   "\n",
  "Group: ",   group_var
))
```

Then pass `tooltip = "text"` to `ggplotly()`. Without that argument, plotly shows
all aesthetics — often cluttered.

**Formatting numbers in the tooltip:**

```r
scales::comma(price)        # → "12,345"
round(rate * 100, 1)        # → 12.3  (combine with paste0 to add %)
scales::dollar(revenue)     # → "$12,345"
```

**Moving the legend below the chart:**

```r
ggplotly(p, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.2))
```


Exercise 4: DT Table {data-icon="fa-table"}
=====================================

Row {data-height=80}
-------------------------------------

### Exercise 4 — Build an interactive summary table with column filters

Dashboards should show **aggregated, summary data** in tables — not individual patient
rows. Build a table where each row is one pathogen × study combination, with prevalence
counts and rates. The user can then filter to a study or search for a pathogen by name.

Row {data-height=480}
-------------------------------------

### Your summary table

```{r}
# ══════════════════════════════════════════════════════════════════════════
# TODO 4 — Compute a summary table and pass it to datatable().
#
# The placeholder below builds a pathogen × study prevalence summary.
# Each row is one pathogen in one study — not an individual case.
#
# Adapt the column selection and formatting to your own data:
#   - Add or remove pathogens from PATHOGENS if using a subset
#   - Swap enr_haz summaries for a continuous variable in your data
#   - Change formatStyle() colors to suit your variable range
# ══════════════════════════════════════════════════════════════════════════
summary_df <- expand.grid(
  Pathogen = pathogen_labels[PATHOGENS],
  Study    = STUDIES,
  stringsAsFactors = FALSE
) %>%
  mutate(
    path_key       = PATHOGENS[match(Pathogen, pathogen_labels)],
    `N cases`      = sapply(Study, function(s) sum(my_data$study == s)),
    `N attr`       = mapply(function(pk, s)
                      sum(my_data[[paste0(pk, "_tac_attr")]] == 1 &
                          my_data$study == s, na.rm = TRUE),
                      path_key, Study),
    `N detected`   = mapply(function(pk, s)
                      sum(my_data[[paste0(pk, "_bin")]] == 1 &
                          my_data$study == s, na.rm = TRUE),
                      path_key, Study),
    `Attr %`       = round(100 * `N attr`     / `N cases`, 1),
    `Detected %`   = round(100 * `N detected` / `N cases`, 1),
    `Mean HAZ`     = round(sapply(Study, function(s)   # TODO: swap for your variable
                      mean(my_data$enr_haz[my_data$study == s], na.rm = TRUE)), 2)
  ) %>%
  select(Study, Pathogen, `N cases`, `N attr`, `Attr %`,
         `N detected`, `Detected %`, `Mean HAZ`)

datatable(
  summary_df,
  options  = list(pageLength = 12, scrollX = TRUE, dom = "ftip"),
  filter   = "top",
  rownames = FALSE,
  class    = "compact stripe"
) %>%
  formatStyle(
    "Attr %",
    background         = styleColorBar(c(0, 30), "#6BAED6"),
    backgroundSize     = "90% 75%",
    backgroundRepeat   = "no-repeat",
    backgroundPosition = "center"
  )
```

### Formatting and dom options

```r
# Round numeric columns
formatRound(columns = "Mean HAZ", digits = 2)

# Format as percentage (multiplies by 100 and adds %)
formatPercentage(columns = "rate_col", digits = 1)

# Inline colour bar — visual encoding without a separate chart
formatStyle(
  "Attr %",
  background         = styleColorBar(c(0, 30), "#6BAED6"),
  backgroundSize     = "90% 75%",
  backgroundRepeat   = "no-repeat",
  backgroundPosition = "center"
)
```

**`dom` string — controls which elements appear around the table:**

| Letter | Element |
|---|---|
| `f` | Global search box |
| `t` | The table |
| `i` | Row count info |
| `p` | Page navigation |
| `l` | "Show N entries" dropdown |
| `B` | Download buttons (add `"B"` to dom and the `Buttons` extension) |

**Note:** In a static dashboard (Part 1), you call `datatable()` directly and it works.
In a Shiny dashboard (Part 2 and this file), wrap it in `renderDT({})` so it reacts to
inputs. Exercise 6 shows this pattern.


Exercise 5: Add Shiny Inputs {data-icon="fa-bolt"}
=====================================

Row {data-height=80}
-------------------------------------

### Exercise 5 — Make your ggplot reactive with Shiny inputs

Add inputs, a `reactive()` to filter the data, and a `renderPlot()` to draw the figure.

Row {data-height=180}
-------------------------------------

### Controls {data-width=300}

```{r}
# ══════════════════════════════════════════════════════════════════════════
# TODO 5a — Replace with inputs that make sense for your data.
# Use prefix ex5_ on all input IDs.
# ══════════════════════════════════════════════════════════════════════════
inputPanel(
  checkboxGroupInput(
    "ex5_group",
    label    = "TODO: filter label",
    choices  = levels(my_data$cyl),       # TODO: your categorical column
    selected = levels(my_data$cyl)
  ),

  sliderInput(
    "ex5_range",
    label = "TODO: numeric filter label",
    min   = floor(min(my_data$mpg)),      # TODO: your numeric column
    max   = ceiling(max(my_data$mpg)),    # TODO: your numeric column
    value = c(floor(min(my_data$mpg)),
              ceiling(max(my_data$mpg)))
  )
)
```

Row {data-height=460}
-------------------------------------

### Reactive plot {data-width=580}

```{r}
# ══════════════════════════════════════════════════════════════════════════
# TODO 5b — Fill in reactive() and renderPlot().
# ══════════════════════════════════════════════════════════════════════════
ex5_data <- reactive({
  my_data %>%
    filter(
      cyl %in% input$ex5_group,          # TODO: your categorical column
      mpg >= input$ex5_range[1],         # TODO: your numeric column
      mpg <= input$ex5_range[2]          # TODO: your numeric column
    )
})

renderPlot({
  d <- ex5_data()
  req(nrow(d) > 0)

  # TODO 5c — Build your ggplot using d (the filtered data).
  ggplot(d, aes(x = wt, y = mpg, color = cyl)) +  # TODO: your variables
    geom_point(size = 2.5, alpha = 0.75) +
    scale_color_brewer(palette = "Set2") +
    labs(
      title    = "TODO: title",
      subtitle = paste0(nrow(d), " records match current filters"),  # keep this
      x        = "TODO: x label",
      y        = "TODO: y label"
    ) +
    theme_minimal(base_size = 13) +
    theme(legend.position  = "bottom",
          panel.grid.minor = element_blank())
}, res = 110)
```

### What to verify {data-width=320}

Once your inputs and plot are working, check these:

**Blank output, not an error, when filter is empty**

If you uncheck all boxes and move the slider to an impossible range, the plot should
show a blank space — not a red error message. That is `req(nrow(d) > 0)` working.
If you see an error instead, make sure `req()` is the second line inside `renderPlot()`.

**Subtitle count updates**

The `paste0(nrow(d), " records match current filters")` line in `subtitle` re-renders
with the plot because `d` and `nrow(d)` are read inside `renderPlot()`.

**Input ID prefix**

All inputs on this tab use `ex5_`. That means they do not interfere with inputs on
other tabs even if they filter the same column.

**Column name mismatch**

The most common error at this stage: the column name in `filter()` does not match the
actual column in `my_data`. Check with `names(my_data)` in the console.


Exercise 6: Shared Reactive {data-icon="fa-link"}
=====================================

Row {data-height=80}
-------------------------------------

### Exercise 6 — Feed one reactive() into both a plot and a table

This exercise demonstrates the most important Shiny pattern: define a `reactive()` once,
then call it from multiple `render*()` functions. The filter runs once; both outputs update.

Row {data-height=180}
-------------------------------------

### Controls {data-width=280}

```{r}
# ══════════════════════════════════════════════════════════════════════════
# TODO 6a — Add inputs. Use prefix ex6_ on all IDs.
# ══════════════════════════════════════════════════════════════════════════
inputPanel(
  checkboxGroupInput(
    "ex6_group",
    label    = "TODO: filter label",
    choices  = levels(my_data$cyl),      # TODO: your categorical column
    selected = levels(my_data$cyl)
  ),

  selectInput(
    "ex6_sort",
    label   = "Sort table by:",
    choices  = numeric_cols,             # defined in global chunk
    selected = numeric_cols[1]
  )
)
```

Row {data-height=420}
-------------------------------------

### Reactive plot {data-width=460}

```{r}
# ══════════════════════════════════════════════════════════════════════════
# TODO 6b — Define ex6_data reactive. Both outputs below will call it.
# ══════════════════════════════════════════════════════════════════════════
ex6_data <- reactive({
  my_data %>%
    filter(cyl %in% input$ex6_group)     # TODO: your column and input
})

renderPlot({
  d <- ex6_data()
  req(nrow(d) > 0)

  # TODO 6c — A summary plot works well here: boxplot or bar chart
  ggplot(d, aes(x = cyl, y = mpg, fill = cyl)) +  # TODO: your variables
    geom_boxplot(alpha = 0.8) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title    = "TODO: title",
      subtitle = paste0(nrow(d), " records"),
      x        = "TODO: x label",
      y        = "TODO: y label"
    ) +
    theme_minimal(base_size = 13) +
    theme(legend.position  = "none",
          panel.grid.minor = element_blank())
}, res = 110)
```

### Reactive table {data-width=440}

```{r}
# ══════════════════════════════════════════════════════════════════════════
# TODO 6d — Use ex6_data() here — the SAME reactive as the plot.
# The filter runs once; both outputs share the result.
# ══════════════════════════════════════════════════════════════════════════
renderDT({
  d <- ex6_data()
  req(nrow(d) > 0)

  d %>%
    arrange(desc(.data[[input$ex6_sort]])) %>%
    datatable(
      options  = list(pageLength = 8, dom = "tip", scrollX = TRUE),
      rownames = FALSE,
      class    = "compact stripe"
    ) %>%
    formatRound(columns = input$ex6_sort, digits = 1)
})
```


Checklist & Next Steps {data-icon="fa-check-square"}
=====================================

Row {data-height=100}
-------------------------------------

### Review your dashboard before sharing it

Work through this checklist. Change `[ ]` to `[x]` in the source `.Rmd` for each item
you have completed.

Row {data-height=380}
-------------------------------------

### Checklist

**Setup**

- [ ] `my_data` loads without errors
- [ ] `numeric_cols` and `categorical_cols` reflect your actual column names
- [ ] The dataset notes box on the Instructions tab is filled in
- [ ] No sensitive or identifying data is included unless intended

**Exercises**

- [ ] Exercise 1: three valueBoxes show meaningful numbers with real labels
- [ ] Exercise 2: static ggplot has a title, labelled axes, and a color or fill aesthetic
- [ ] Exercise 3: ggplotly tooltip shows at least two useful columns from the data
- [ ] Exercise 4: DT table has column filters and at least one format function applied
- [ ] Exercise 5: reactive plot updates on input change; shows blank (not error) when filter is empty
- [ ] Exercise 6: plot and table both call the same `reactive()` and update together

**Polish**

- [ ] All `# TODO` comments replaced with real code
- [ ] All `"TODO: ..."` strings replaced with real labels and titles
- [ ] All placeholder text in Markdown boxes updated
- [ ] Input IDs use per-tab prefixes (`ex5_`, `ex6_`, etc.)

### Stretch goals

**Layout**

- Add a new page tab for a second view of your data
- Use `{.tabset}` on a row to create sub-tabs within a page
- Add a `{.sidebar}` column to one page (see Example 1 in Part 2)
- Try switching `orientation: columns` in the YAML

**Outputs**

- Add a `gauge()` showing a rate or percentage
- Add download buttons to the DT table: set `dom = "Bftip"` and add
  `extensions = "Buttons"` to `datatable()`
- Add `facet_wrap()` to the static ggplot

**Shiny**

- Add `renderText()` to show a dynamic caption below a plot
- Add a `radioButtons()` that switches the y-axis variable in the reactive plot
- Add `conditionalPanel()` to show an extra input only when a specific choice is selected

Row {data-height=220}
-------------------------------------

### Sharing your finished dashboard

**As a static HTML file** — remove `runtime: shiny` from the YAML, remove all Shiny input
and render functions (replace `renderPlot({...})` with plain ggplot calls), then click
**Knit**. The result is a single `.html` file you can email.

**As a live Shiny dashboard** — deploy to shinyapps.io:

```r
install.packages("rsconnect")
rsconnect::setAccountInfo(name   = "your-account",
                          token  = "YOUR_TOKEN",
                          secret = "YOUR_SECRET")
rsconnect::deployApp()   # run from the folder containing your .Rmd
```

After deploying you receive a public URL. The free tier on shinyapps.io allows
25 active hours per month. For team or institutional use, Posit Connect provides
authentication, scheduled refreshes, and usage monitoring.

